Safe Memory Regions for Big Data Processing

### 2016-10-19 paper/implementation.tex
implementation

We have implemented the prototype of \name compiler frontend,
including its region type system and type inference, in 3k+ lines of
OCaml. Our implementation is called \namec. The input to \namec is a
program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
includes assignments, conditionals, loops, more primitive datatypes
(\eg, integers), and a null value. 
% If needed, a specification file containing region type annotations
% for higher-order functions can also be provided. 
Our implementation of region type inference and constraint solving
closely follows the description given in
Sec.~\ref{sec:type-inference}.

% To solve the
% constraints that arise during type inference, we built a solver called
% \csolve that implements constraint solving approach based on fixpoint
% computation and graph augmentation described in \S~\ref{sec:csolve}. 
% Since graph augmentation is NP-hard, \csolve uses an approximation
% algorithm that repeatedly adds an edge and recomputes paths in graph
% $G_1$ until it is as connected as graph $G_2$. Though this approach
% may not necessarily return the weakest solution to the abduction
% problem (\S~\ref{sec:csolve}), it seems to do so in practice, since
% constraints are not often complex. 
% If the input $\absof{\FB^+}$ program does not create any unsafe
% references, \namec annotates it with region types, which act as a
% witness to program's memory safety. If potential safety violations are
% encountered, then the type inference fails during the constraint
% solving phase.

@@

To evaluate the practical utility of our region type system and type
inference, we 
% implemented some of the C\# libraries in $\absof{\FB^+}$, and 
performed two kinds of experiments. First, we implemented some of the
microbenchmarks ($\le$100 LOC), which are the standard libraries such
as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
our inference engine to infer their principal region types. Since a
library class is region-oblivious, if it is well-typed as per the core
type system, then \namec must be able to automatically construct its
region-type-annotated definition without fail. As expected, \namec was
able to infer principal region types for all the library classes,
under 10ms. Fig~\ref{fig:rev} shows the region-type-annotated
definition computed for the list reverse method. Observe that \namec
was able to infer that the list and its data (of type \C{T}) can be
allocated in different regions, as long as the later outlives the
former. This allows, for instance, a \C{preOrder} method to traverse a
tree in a transferable region, and return a list of its nodes, where
the list itself is allocated in the stack region.

Next, we translated some of the Naiad streaming query operator
benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
$\absof{\FB^+}$, and used \namec to verify their safety. During the
process, we found multiple instances of potential memory safety
violations in the $\absof{FB^+}$ translation of benchmarks, which we
verified to be present in the original C\# implementation as well. The
cause of all safety violations is the creation of a reference from the
outgoing message (a transferable region) to the payload of the
incoming message. For example, the implementation of \C{SelectVertex}
contains the following:
\begin{codejava}
  if (this.selector(inMsg.payload[i])) {
    outMsg.set(outputOffset, inMsg.payload[i]);
    ...
  } 
\end{codejava}
The \C{outMsg} is later transferred to a downstream actor, where the
reference to \C{inMsg}'s payload becomes unsafe\footnote{This unsafe
reference could have gone unnoticed during experiments
in~\cite{Broom:HotOS} because their experimental setup included only
one actor.}. We eliminated such unsafe references by creating a clone
of \C{inMsg.payload[i]} in \C{outMsg}, and our compiler was
subsequently able to certify the safety of all references.

Our experience with Naiad benchmarks shows the utility of our
type inference/checking tool,
%suggests that, although our
%approach cannot statically enforce memory safety, it is nonetheless
%useful, 
particularly because it comes at no additional cost to the developer.
If the developer is satisfied with
the confidence gained by static verification, she may even choose to
turn the runtime safety checks off. Since we did not implement code
generation in \namec, we could not measure the runtime overhead of
checks needed to enforce safety. However, the number of LOC performing
operations on \C{Region} objects relative to the total LOC never
exceeds 8\% in Naiad benchmarks. 
% \C{SelectVertex}, for example, contains one each of
% \C{open} and \C{transfer} operations on one region in its 45LOC,
% whereas, \C{JoinVertex} contains 5 \C{open}s, 2 \C{transfer}s, and 1
% \C{free} operation.

% \paragraph{Memory safety violation}

>>>>

We have implemented \namec, a prototype of \name compiler frontend,
including its region type system and type inference, in 3k+ lines of
OCaml.
% Our implementation is called \namec.
The input to \namec is a program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
includes assignments, conditionals, loops, more primitive datatypes
(\eg, integers), and a null value. 
% If needed, a specification file containing region type annotations
% for higher-order functions can also be provided. 
Our implementation of region type inference and constraint solving
closely follows the description given in
Sec.~\ref{sec:type-inference}.

We performed two kinds of experiments to evaluate our region type system and type
inference.
First, we implemented some 
microbenchmarks ($\le$100 LOC) consisting of standard classes such
as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
our inference engine to infer their region types.
These classes are region-oblivious. Hence, as long as they are well-typed
as per the core type system, \namec must be able to automatically construct its
region-type-annotated definition without fail. \namec was
able to infer the expected region types for all these classes under 10ms.
Fig~\ref{fig:rev} shows the region-type-annotated
definition computed for the list reverse method. Observe that \namec
was able to infer that the list and its data (of type \C{T}) can be
allocated in different regions, as long as the later outlives the
former. This allows, for instance, a \C{preOrder} method to traverse a
tree in a transferable region, and return a list of its nodes, where
the list itself is allocated in the stack region.

@@

Next, we translated some of the Naiad streaming query operator
benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
$\absof{\FB^+}$, and used \namec to verify their safety.
The number of LOC performing operations on \C{Region} objects relative
to the total LOC is 8\% or under in the Naiad benchmarks. 
During the process, we found multiple instances of potential memory safety
violations in the $\absof{FB^+}$ translation of benchmarks, which we
verified to be present in the original C\# implementation as well. The
cause of all safety violations is the creation of a reference from the
outgoing message (a transferable region) to the payload of the
incoming message. For example, the implementation of \C{SelectVertex}
contains the following:
\begin{codejava}
  if (this.selector(inMsg.payload[i])) {
    outMsg.set(outputOffset, inMsg.payload[i]);
    ...
  } 
\end{codejava}
The \C{outMsg} is later transferred to a downstream actor, where the
reference to \C{inMsg}'s payload becomes unsafe\footnote{This unsafe
reference could have gone unnoticed during experiments
in~\cite{Broom:HotOS} because their experimental setup included only
one actor.}. We eliminated such unsafe references by creating a clone
of \C{inMsg.payload[i]} in \C{outMsg}, and our compiler was
subsequently able to certify the safety of all references.

Our experience with Naiad benchmarks shows the utility of our
type inference/checking tool,
particularly because it comes at no additional cost to the developer.
%% If the developer is satisfied with
%% the confidence gained by static verification, she may even choose to
%% turn the runtime safety checks off. Since we did not implement code
%% generation in \namec, we could not measure the runtime overhead of
%% checks needed to enforce safety.

% \C{SelectVertex}, for example, contains one each of
% \C{open} and \C{transfer} operations on one region in its 45LOC,
% whereas, \C{JoinVertex} contains 5 \C{open}s, 2 \C{transfer}s, and 1
% \C{free} operation.
